---
title: 'Class 10: Unsupervised Learning Mini Project'
author: "Preyas Soni"
date: "2/6/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of Human Breast Cancer Cells 

First we need to import our input the data, it was an online link 

```{r}
wisc.df <- read.csv("https://bioboot.github.io/bimm143_W18/class-material/WisconsinCancer.csv")
head(wisc.df)
```
```{r}
# Convert the features of the data: wisc.data, there are some funky things in this dataset that we will ignore for our analysis. this includes our ID and diagnosis coulum and the funny last x couloumn
wisc.data <- as.matrix( wisc.df[,3:32])
head(wisc.data)
```

Q1. How many patients do we have data for? 

```{r}
nrow(wisc.data)
```

Q2. How many cancer and non cancer? 

```{r}
table(wisc.df$diagnosis)
```

Q3. how many variables in the data are suffixed with _ mean 

```{r}
colnames(wisc.data)
```
```{r}
grep("_mean", colnames(wisc.data), value = TRUE)
```

```{r}
length(grep("_mean", colnames(wisc.data)))
```

## Principal Component Analysis 

Before we turn to PCA we need to think, whether we should SCALE our input. 

Recall two common reasons for scaling data include:

The input variables use different units of measurement.
The input variables have significantly different variances.
Check the mean and standard deviation of the features (i.e. columns) of the wisc.data to determine if the data should be scaled. Use the `colMeans()` and `apply()` functions like you’ve done before.

```{r}
round(apply(wisc.data, 2, sd ), 2 )
```


Looks like we need to set scale=TRUE! 




```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp( wisc.data, scale=TRUE )
summary(wisc.pr)
```

Q4. From your results, what proportion of the original variance is captured by the principal components 

The 1st PC captures 44.27% of the original variance. Nore that 72.6% is captured in the first 3 PCs.. 

Lets make some figures... 

```{r}
biplot(wisc.pr)
```

OOF! THis is a hot mess 

We need to do our own PC1 vs PC2 plot and lets color by the diagnosis 

```{r}
attributes(wisc.pr)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=wisc.df$diagnosis, xlab= )
abline(h=0, col="gray", lty=2)
abline(v=0, col="gray", lty=2)
```

CLuster in PC space 

First lets see if we can cluster the oiginal data 

```{r}
hclust( dist(wisc.data) )
hclust
```


Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with complete linkage. Assign the results to wisc.pr.hclust.
```{r}
wisc.pr.hclust <- hclust( dist(wisc.pr$x[,1:3]), method="ward.D2")
plot(wisc.pr.hclust)
cutree(wisc.pr.hclust, k=3)
```
```{r}
grps3 <- cutree(wisc.pr.hclust, k=3)
table(grps3)
```
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=grps3)
```

```{r}
table(grps3, wisc.df$diagnosis)
```

```{r}

## we will use predict function 

url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata = new)
npc
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=wisc.df$diagnosis)
points(npc[,1], npc[,2], col="purple", pch=15, cex=3)
text(npc[,1], npc[,2], labels = c(1,2), col="yellow")
```

